---
title: "Geriatric Depression Scale"
subtitle: "Original version: Yesavage J A, 1983 - Norwegian version: Knut Engedal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(knitr)
library(kableExtra)
devtools::load_all(".")
```


## Background

The Geriatric Depression Scale (GDS) is an instrument designed specifically for rating depression in the elderly. 
It can be administrated to healthy, medically ill, and mild to moderately cognitively impaired older adults. 
As a general rule, GDS is administrated in LCBC to older adults with a lower cut off around 60 years. 
However, please consult the instructions for each project, as this guideline has been implemented at different time points across the projects. 

The questionnaire consists of 30 questions tapping into a wide variety of topics relevant to depression, including cognitive complaints, motivation, thoughts about the past and the future, self-image, and mood itself. 
The answers should be based the participants' feelings throughout the last week. 

Twenty of the questions indicate the presence of depression when answered positively, while the ten remaining indicate depression when answered negatively (see scoring instructions below). 
The questionnaire is scored accordingly, giving one point for each statement that affirms a depressive symptom. 
The sum of these scores yields one total score, with a possible range between 0 and 30.


## GDS scoring information
The GDS is quite straight forward in its format, a series of 30 questions that take a yes or no answer. 
This binary coding makes it quite easy to work with.
Several of the questions, however, are formulated in such a way that they require a reversal of the coding before the total score can be summed.
The questions which require reversal of coding are, **01, 05, 07, 09, 15, 19, 21, 27, 29, 30**, meaning answering "yes" to these should be altered to 0, and "no" altered to 1, before calculating the sum score.
The total GDS score is after reversal, a simple addition of all the answers into a single score.

One point is given for any “No” answered to the following questions:  
1, 5, 7, 9, 15, 19, 21, 27, 29 and 30

and one point is given for every “Yes” answered on the following questions:  
2, 3, 4, 6, 8, 10, 11, 12, 13, 14, 16, 17, 18, 20, 22, 23, 24, 25, 26, 28  

### Depression categories
There are 3 categories of severity for the GDS total score.
Below or equal to `9` is "Normal", above 19 is "Severe depression", and the remaining fall within "Mild depression". 

|GDS score|Depression category|
|---|---|
|0-9|Normal|
|10-19|Mild depressive|
|20-30|Severe depressive|	      

## Data requirements
### Column naming
The easiest is to have data coded as in the MOAS, as this will let you use default values for the arguments. 
The column names in the MOAS all start with `GDS_` and then are followed by a two-digit numbering of the question: 

GDS_01,  
GDS_02,   
GDS_03,   
...  
GDS_28,  
GDS_29,   
GDS_30  

If your data is coded differently, a consistent naming scheme should help you use the functions anyway.

### Data values
Each row of data should belong to a single answer to the entire questionnaire.
Meaning if you have multiple answers to the questionnaire over time, these should be placed in another row, duplicating the participant ID, together with a column indicating the timepoint the data was collected in.
Data values are binary yes and no answers to the GDS. 
While the functions are made in such a way that any type of binary coding works well, the default is set to be `yes = 1`, `no = 0`.
These can be altered by applying the `gds_values` functions to the other functions asking for the coding schema.

```{r}
gds_values(yes = 1, no = 0)
gds_values(yes = "yes", no = "no")
gds_values(yes = "ja", no = "nei")
```


## Using the `gds` functions

```{r}
library(dplyr)
library(tidyr)
library(stringr)

# randomly create some data as example
gds_data <- expand_grid(ID = 1:5, 
            key = "GDS", 
            question = str_pad(1:30, 2, "left", 0)) %>% 
    mutate(value = sample(c(0,1), nrow(.), replace = TRUE)) %>% 
  unite(key, c(key, question)) %>% 
  spread(key, value)

gds_data %>% 
  # ignore this part, it's just so the table looks better online
  kable() %>% 
  kable_styling()
```

### Calculating the GSD sum
The function to calculate the total GDS score is `gds_compute_sum`, and requires a data frame, a tidy-selector that selects all the GDS columns and a tidy-selector indicating which columns should be reversed.
If you have MOAS-like data, then the defaults should work, and as such you need not specify this.

```{r}
gds_data %>% 
  gds_compute_sum()
```

To add a column with total GDS to a data set, place the argument inside a `mutate`, here a column named `GDS_total` will appear as the right-most column.

```{r}
gds_data %>% 
  mutate(
    GSD_total = gds_compute_sum(.)
  ) %>% 
  # ignore this part, it's just so the table looks better online
  kable() %>% 
  kable_styling()
```

### Turning the sum into a factor
Most often, people will report not only the actual score, but also the accompanying depression classification based on that score.
This can be derived by using the `gds_factorise` function.


```{r}
gds_data %>% 
  gds_compute_sum() %>% 
  gds_factorise()
```

and this can also be added directly to the data through a `mutate``

```{r}
gds_data %>% 
  mutate(
    GSD_total = gds_compute_sum(.),
    GDS_cat =  gds_factorise(GSD_total)
  ) %>% 
  # ignore this part, it's just so the table looks better online
  kable() %>% 
  kable_styling()
```

### Computing score and factor in one function
Lastly, since it is most common to want both the score and the factor added directly do data, a convenience function exists to assist in that.

```{r}
gds_data %>% 
  gds_compute()
```

alternatively, you can also `keep_all = FALSE` in that function, to only retain the two computed columns, while maintaining the remaning data structure.

```{r}
gds_data %>% 
  gds_compute(keep_all = FALSE)
```



## References

J A Yesavage, T L Brink, T L Rose, O Lum, V Huang, M Adey, V O Leirer (1982), Development and Validation of a Geriatric Depression Screening Scale: A Preliminary Report, _J Psychiatr Res_, 17 (1), 37-49, [doi: 10.1016/0022-3956(82)90033-4](https://www.sciencedirect.com/science/article/abs/pii/0022395682900334?via%3Dihub)

E L Lesher 1, J S Berryhill (1994), Validation of the Geriatric Depression Scale -- Short Form Among Inpatients, _J Clin Psychol_, 50 (2), 256-60,
[doi: 10.1002/1097-4679(199403)50:2<256::aid-jclp2270500218>3.0.co;2-e](https://doi.org/10.1002/1097-4679(199403)50:2%3C256::aid-jclp2270500218%3E3.0.co;2-e)
