---
title: "International Physical Activity Questionnaire (IPAQ)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{International Physical Activity Questionnaire}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)

```

## Background

The purpose of the International Physical Activity Questionnaires (IPAQ) is to provide a set of well-developed instruments that can be used internationally to obtain comparable estimates of physical activity. There are two versions of the questionnaire. The short version is suitable for use in national and regional surveillance systems and the long version provide more detailed information often required in research work or for evaluation purposes.

## Rationale

The public health burden of a sedentary lifestyle has been recognized globally, but until recently, the prevalence and impact of the problem has not been studied in a uniform and systematic fashion. The questionnaire is the most feasible instrument for measuring physical activity in large groups or populations. However, many of the existing instruments are not comparable in the type of activities surveyed (i.e., leisure-time activities only) and format for data collection. 

In response to the global demand for comparable and valid measures of physical activity within and between countries, IPAQ was developed for surveillance activities and to guide policy development related to health-enhancing physical activity across various life domains.

## Scoring

Scoring of the IPAQ is based on a metric called METs, which are multiples of the resting metabolic rate.
[The IPAQ scoring description can be found here ](https://docs.google.com/viewer?a=v&pid=sites&srcid=ZGVmYXVsdGRvbWFpbnx0aGVpcGFxfGd4OjE0NDgxMDk3NDU1YWRlZTM)


### Continuous Score

Expressed as MET-min per week: MET level x minutes of activity/day x days per week

MET levels:

* **Light** - 3.3 METs

* **Moderate** - 4.0 METs

* **Vigorous** - 8.0 METs

Total MET-minutes/week = Light (3.3 x min x days) + Mod (4.0 x min x days) + Vig (8.0 x min x days)


### Categorical Score

Three levels (categories) of physical activity are proposed:

#### Category 1: Low

This is the lowest level of physical activity. Those individuals who not meet criteria for categories 2 or 3 are considered low/inactive. 

#### Category 2: Moderate

Any one of the following 3 criteria:

* 3 or more days of vigorous activity of at least 20 minutes per day **OR**

* 5 or more days of moderate-intensity activity or walking of at least 30 minutes per day **OR**

* 5 or more days of any combination of walking, moderate-intensity or vigorous intensity activities achieving a minimum of at least 600 MET-min/week. 


#### Category 3: High

Any one of the following 2 criteria:

* Vigorous-intensity activity on at least 3 days and accumulating at least 1500 MET-minutes/
week **OR**

* 7 or more days of any combination of walking, moderate-intensity or vigorous
intensity activities achieving a minimum of at least 3000 MET-minutes/week 

## Using the `ipaq` functions

### Altering times
First important notice is on how your time-data is coded.
The IPAQ has three important questions on the number of minutes spent doing something. 
This can be recorded in many ways, and some might have given participants options to answer in specific ways to reduce inaccurate data.
In LCBC we have online solutions for data collection, and for time questions we have forced an `HH:MM` format to make sure respondents are consistent in the way they answer. 
In order to use these with the remaining ipaq calculations, a conversion to minutes is necessary.
The `ipaq_time_atler` function will alter the columns from `HH:MM` to minutes.
If you have any `NA` in the time data, you will get a warning about not being able to parse the data,
this is expected and wanted behaviour, but you may proceed, the METs will not be calculated where data is missing.

```{r}
library(Questionnaires)
library(dplyr)

dat <- data.frame(
  time_1  = c("02:34", "09:33", "01:14"),
  time_2  = c("00:55", NA, "00:30")
)

dat

ipaq_time_alter(dat, cols = c(time_1, time_2))
```

### MET category scores
To calculate the MET score for an individual MET category the `ipaq_compute_met` is what you need.
This function takes a vector of minutes, days and the MET-factor you want to apply. 
Default MET-factor can be found in the `ipaq_mets()` function.

```{r}
ipaq_mets()
```

The different MET categories have different questions in the IPAQ, and should be calculated separately before combined to the total score. 

```{r}
vig_data <- data.frame(
 ipaq_vig_mins = c(60, 20, 60, 25, 90, 20, 0, 75, 60, 30),
 ipaq_vig_days = c(1, 3, 2, 5, 6, 1, 1, 2, 2, 4)
)

vig_data

ipaq_compute_met(vig_data$ipaq_vig_mins, 
                 vig_data$ipaq_vig_days, 
                 met = 8.0)
```


The total MET score can be manually computed using the `ipaq_compute_sum`, where one supplies a vector for each of the three MET categories, containing the pre-calculates MET scores for the categories.
```{r}
light = c(1300, 300)
moderate = c(200, 400)
vigurous = c(0, 1300)

ipaq_compute_sum(vigurous , moderate, light)
```

Finally, you can calculate everything in one go using the `ipaq_compute` function.
This function has many arguments, as the minutes and days of each category (2x3) must
be specified, and also the mets you want applied must be supplied.
By default, if you have MOAS-like data you need not do any extra specifications to make this work,
the MOAS columns necessary are specified as defaults.
The easiest way to alter the default mets is by using the `ipaq_mets` function to alter them.
The function takes an entire data frame with all columns of data necessary to compute the MET score.

```{r}

data <- data.frame(
  ipaq_1a = c(NA, "No", "Yes", NA, NA, NA, NA, NA, NA, "Yes"), 
  ipaq_1b = c(0, NA, 3, 3, 0, 2, 4, 3, 0, 3), 
  ipaq_2 = c("00:00", "", "01:00", "01:00", "00:00", "03:30", "01:00", "00:25", "00:00", "00:45"), 
  ipaq_3a = c(NA, "Yes", "Yes", NA, NA, NA, NA, NA, NA, "No"), 
  ipaq_3b = c(1, 3, 1, 3, 1, 4, 0, 5, 4, NA), 
  ipaq_4 = c("00:30", "01:30", "01:00", "01:00", "01:00", "02:00", "00:00", "00:15", "03:00", ""), 
  ipaq_5a = c(NA, "Yes", "Yes", NA, NA, NA, NA, NA, NA, "Yes"), 
  ipaq_5b = c(7, 3, 7, 7, 3, 3, 0, 5, 7, 4), 
  ipaq_6 = c("01:00", "00:20", "01:00", "00:25", "01:30", "00:20", "00:00", "01:15",  "01:00", "00:30"), 
  ipaq_7 = c("05:00", "12:00", "05:00", "07:00", "00:18", "05:00", "00:00", "08:00", "04:00", "08:00"), 
  ipaq_8a = c(5, NA, NA, 4, 4, 6, 5, 4, 6, NA),
  ipaq_8b = c(5, NA, NA, 4, NA, 6, 5, 4, 6, NA), 
  ipaq_8c = c(5, NA, NA, 5, 3, 5, 5, 3, 6, NA)
)

data

data %>% 
  ipaq_time_alter() %>% 
  # Keeping only calculated values for show, you like want this to be TRUE (default)
  ipaq_compute(keep_all = FALSE)
```


## References

Ainsworth, B. E., et al. (2000). "Comparison of three methods for measuring the time spent in physical activity." [*Med Sci Sports Exerc* 32: S457-64](https://www.ncbi.nlm.nih.gov/pubmed/10993415)

Ainsworth, B. E., et al. (2006). "Comparison of the 2001 BRFSS and the IPAQ Physical Activity Questionnaires." [*Med Sci Sports Exerc* 38: 1584-92](https://www.ncbi.nlm.nih.gov/pubmed/16960519)

Brown, W. J., S. G. Trost, A. Bauman, K. Mummery and N. Owen (2004). "Test-retest reliability of four physical activity measures used in population surveys." [*J Sci Med Sport* 7: 205-15](https://www.ncbi.nlm.nih.gov/pubmed/15362316)

Hallal, P. C. and C. G. Victora (2004). "Reliability and validity of the International Physical Activity Questionnaire (IPAQ)." [*Med Sci Sports Exerc* 36: 556](https://www.ncbi.nlm.nih.gov/pubmed/15076800)

Mader, U., B. W. Martin, Y. Schutz and B. Marti (2006). "Validity of four short physical activity questionnaires in middle-aged persons." [*Med Sci Sports Exerc* 38: 1255-66](https://www.ncbi.nlm.nih.gov/pubmed/16826022)


